---
title: "Functions, Conditionals, Loops, Tibbles, and Tidy Data"
author: "Bonfire-113"
date: 2023-03-31
---

## Functions and Conditionals

### Functions

#### When should we use functions?

When I have to type a line of code multiple times to accomplish the same thing.

WHY?

Reproducibility/Automation

Once I build a function, I can use across my programs

```{r}
our_function <- function(optional_parameters){
  #CODE
}
our_function
```

An example:

```{r}
# Create a function that can convert C to F
# Formula for conversion is C*(9/5) + 32

test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 21.976820, 21.173577, 21.048836, 18.769735)

# Then we create a function
c_to_f <- function(v){
  temp_fahr <- v*(9/5) + 32
  
  return(temp_fahr)
}
c_to_f(26)
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
c_to_f(test_temp)
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow=2)
test_mat
# Apply:
# First, we gave it the X, or Data source
# Then we told it where to apply the function to, 1 being rows, 2 being columns
# Then, we told it what function to apply. Sum is a built-in R function
apply(test_mat, 1, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 21.976820, 21.173577, 21.048836, 18.769735), nrow = 2)

test_temp_mat
apply(test_temp_mat, 1, c_to_f)
```

#### More Functions!

Now, let's create a function to check if a number is even or odd!

```{r}
check_num <-  function(x){
  if (x %% 2 == 0){
    print('Even')
  } else {
    print('Odd')
  }
}

check_num(4)
check_num(7)
```

Next, let's calculate a factorial:

```{r}
calc_factorial <- function(x){
  if(x < 0){
    print('Factorial is not defined or negative numbers.')
  }else if(x==0){
    print('Factorial of 0 is 1')
  }else {
    result <- 1
    for(i in 1:x){
    result <- result * i
    }
    # Paste is a concat function. Same as F string
    print(paste('Factorial of', x, 'is', result, 'x'))
  }
}

calc_factorial(5)
calc_factorial(-3)
calc_factorial(0)
```

What if we wanted to use a function on a vector?:

```{r}
# Calc the mean of a vector
# Aggregate a vector into a single number
calc_mean <- function(x){
  # Is it empty?
  if(length(x) == 0){
    print('Cannot calculate the mean of an empty vector.')
  }else{
    sum <- 0
    for(i in 1:length(x)){
      sum <- sum + x[i]
    }
    print(paste0('Mean of vector is: ', sum/length(x), '.'))
  }
}

calc_mean(c(2,4,6))
calc_mean(c(1,3,5,7))
```

```{r}
# Max of vector
calc_max <- function(x){
   if(length(x) == 0){
    print('Cannot calculate the mean of an empty vector.')
  }else{
    max <- x[1]
    for(i in 2:length(x)){
      if(x[i] > max){
        max <- x[i]
      }
    }
    print(paste0('Maximum value of the vector is: ', max, '.'))
  }
}

calc_max(C(2,4,6))
```

## In Class Excercise 1:

Create a function that checks if a number is positive or negative. Your function should account for zero.

Test it using a positive and negative number as well as 0.

```{r}
calc_pos_or_neg <- function(x){
  if(x<0){
    print('Negative number')
  }else if(x==0){
    print('Number is 0')
  }else{
    print('Positive number')
  }
}

calc_pos_or_neg(1)
calc_pos_or_neg(0)
calc_pos_or_neg(-2)
```

## Conditionals

The basic syntax for an if statement goes like this:

```{r}
# In R, we follow JavaScript
#if(logical_statement){
  #do something
#}else{
  #do something else
#}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10

if(a<=5){
  print(a)
}else{
  print('its more than 5')
}
```

We can also add else if statements in the middle if we need more options.

```{r}
# In python, we write this as elif
# In R, we use else if

b <- 15

if(b<=5){
  print('small number')
} else if(b > 5 & b<=10){
  print('medium number')
} else{
  print('big number')
}
  
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
c <- 5

if(c/2 != 0){
  if(c==5){
    if(c < 10){
    print('c is 5')
    }  
  }
}
```

Nested ifelse statement: This is a powerful way to evaluate multiple conditions within a single ifelse statement. The syntax for nested ifelse statements is ifelse(condition1, value1, ifelse(condition2, value2, ifelse(condition3, value3, value4))). Here is an example:

```{r}
x <- 10
y <- ifelse(x < 5, 'low', ifelse(x<10, 'medium', 'high'))
y
```

Switch statement: The switch statement is similar to a series of ifelse statements but can be more efficient and easier to read. The syntax is switch(expression, value1, value2, ..., valueN).

```{r}
z <- 2
switch(z, 'one', 'two', 'three')
```

Case_when statement: The case_when statement is a versatile way to evaluate multiple conditions and assign values. The syntax is case_when(condition1 \~ value1, condition2 \~ value2, ..., TRUE \~ valueN).

```{r}
x <- 10
case_when(x < 5 ~ 'low', x < 10 ~ 'medium', TRUE ~ 'high')
```

Short-circuit evaluation: Short-circuit evaluation is a way to evaluate conditional statements more efficiently by only checking the second condition if the first condition is true. The syntax is condition1 && condition2.

```{r}
x <- 10
y <- if(x < 5 || x > 20){
  print('out of range')
} else{
  print('in range')
}
```

## In Class Exercise 2:

Create two vectors of numbers. Using a conditional statement, change the value for the vectors into a character type:

```{r}
v1 <- 1
v2 <- 4

switch(v1, 'one', 'two', 'four')
case_when(v2 < 5 ~ 'small', v2 < 10 ~ 'middle')
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for(val in x){
  ####
}
```

```{r}
for (value in 1:5){
  print(paste(value, 'is greater than', value -1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
vec <- c(1,1,1,1,1,1,1,1)

# Change the first, third, and sixth elements in my vector the value of 6
for(i in c(1,3,6)){
  vec[i] <- 6
}
vec
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow=5, ncol=5)

for(row in 1:3){
  for(col in 3:5){
    mat[row,col] <- 15
  }
}
mat
```

```{r}
mat <- matrix(1:12, ncol=4)
mat
means <- numeric(nrow(mat))

for(i in 1:nrow(mat)){
  means[i] <- mean(mat[i,])
}
print(means)
```

```{r}
mat <- matrix(1:12, ncol = 4)
sums <- numeric(ncol(mat))

for(i in 1:ncol(mat)){
  sums[i] <- sum(mat[,i])
}
sums
```

Like Python, we have while loops in R:

```{r}

```

## In Class Exercise 3:

Write a while loop that calculates the sum of the numbers 1-10:

```{r}

```

## Tibbles

### What is a Tibble?

A tibble is a dataframe that has a "modern" take. They are designed to make working with data easier and more intuitive. It is part of our `tidyverse` package!!!!

```{r}
library(tidyverse)
```

```{r}
# Create a tibble with different data types

num <- c(1,2,3)
char <- c('John','Jane','Jimmy')
fact <- factor(c('male', 'female', 'male'))
logi <- c(TRUE, FALSE, TRUE)

my_tbl <- tibble(num, char, fact, logi)
my_tbl
```

#### Data Frames VS Tibbles

Some main differences between the two:

-   Tibbles print only the first few rows and all columns that fit on the screen, making it easier to view large data sets without overwhelming the console.

-   Tibbles also show their data type for each column (for example: `<chr>` for character, `<dbl>` for doubles)

-   Tibbles have a stricter subsetting mechanism than data frames. When you extract a single column from a tibble, it returns a tibble instead of a vector. When you extract a single row, it returns a tibble with one row instead of a data frame with one row. This behavior helps avoid common programming errors.

-   Tibbles preserve the case of variable names, while data frames do not. Tibbles also allow column names to be specified with backticks, which is useful when using reserved words or names with spaces or special characters.

-   Tibbles are more consistent than data frames in their treatment of data types. They do not coerce character vectors to factors, and they do not silently drop dimensions when subsetting.

-   Tibbles treat missing values more consistently than data frames. Missing values are always displayed as "NA", regardless of the data type. This avoids the confusion that can arise when missing values are represented differently for different data types (e.g., "NA", "NaN", ".")

```{r}
my_df <- data.frame(num,char,fact,logi)
my_df

my_tbl <- tibble(num, char, fact, logi)
my_tbl
```

We can see that tibbles offer us much more information! They are also easier to manipulate, work well with tidyr and dplyr.

What if we wanted to convert a Data Frame object into a Tibbles object? We have a method for that!

```{r}
df <- data.frame(x=1:5, y=letters[1:5], z=c(TRUE, FALSE,TRUE,FALSE,TRUE))
tb <- as_tibble(df)
```

If we need to make a tibble display more columns than it thinks it should, we can override it with `print()`:

```{r}
print(tb)
```

The best way to see more of the data than a tibble default displays is to use View(). Nice!

```{r}
view(tb)
```

#### Subsetting

```{r}
df$x
tb$x
```

At this point, it should be pointed out that the transformation functions from `dplyr` are extremely valuable in context of subsets. In most programming, there are multiple ways of finding the same answer; this applies here as well.

You can also turn a tibble back into a data frame with `as.data.frame()`. This only really comes into play if you're interacting with older code.

## Reading Data

```{r}
view(read_csv('C:/Users/Noki/Downloads/femaleMiceWeights.csv'))
```

## Tibble Functions

Tibble and dplyr work wonders together! We can manipulate Tibbles using dplyr functions, which provide us a streamlined way to filter, select, group, summarize, and mutate data! \### `filter()`

```{r}
filter(my_tbl, num > 1)
```

### `select()`

```{r}
select(my_tbl)
```

### `group_by()`, `summarize()`

```{r}
group_by(my_tbl, fact) %>% summarize(avg_num = mean(num))
```

### `mutate()`

```{r}
mutate(my_tbl, num_squared = num^2)
```

## N/A Values

Tibbles handle missing data in a similar way to data frames. By default, missing values are represented by NA (not available) in tibbles. However, there are several functions available in R to handle missing values in tibbles.

It's worth noting that tibbles also have a special column type for missing values called missing. This can be useful in cases where NA is a valid value in a column, but we want to be able to distinguish between missing values and non-missing values.

One of the most common functions used to handle missing values in tibbles is na_if(), which can be used to replace specified values with NA.

```{r}
tb <- tibble(x=c(1,2,'missing',4,5,'missing'))
tb
tb <- tb%>%
  mutate(x=na_if(x,'missing'))
tb
```

Now that we replaced all our "missing" values with a NA, let's explore how tibbles handles replacing these values!

Tibbles uses the function `replace_na()`, which we can use to replace all the `NA` values with a specified value.

```{r}
tb <- tb %>%
  mutate(y = replace_na(y, 0))
tb
```

## Importing/Exporting Data

Tibbles can be imported and exported using a variety of file formats! Some examples of this are CSV, Excel, and even SQL databases!

```{r}
my_data <- read_csv('C:/Users/Noki/Downloads/femaleMiceWeights.csv') %>% as_tibble()
my_data

# Mutation
# HERE

# Save to new file
write_csv(my_data, 'my_data.csv')
```

## Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    # 1. 
    # You will be given an array a and a value x. All you need to do is check whether the provided array contains the value.

    # Array can contain numbers or strings. X can be either.
    # Return true if the array contains the value, false if not.

    check <- function(a, x){
      return(any(a == x))
    }

    ```

    ```{r}
    # 2.
    # Write a function to split a string and convert it into an array of words.
    string_to_array <- function(s){
      unlist(strsplit(s, ' '))
    }

    s = "I love arrays they are my favorite"
    string_to_array(s)
    ```

    ```{r}
    # 3.
    # It's pretty straightforward. Your goal is to create a function that removes the first and last characters of a string. You're given one parameter, the original string. You don't have to worry with strings with less than two characters.

    removeChar <- function(str){
      str_sub(str,2,-2)
    }

    removeChar('eloquent') # Getting errors on CodeWars. 
    # removeChar <- function(str){
    # Correct solution from codewars:
    #substr(str,2, nchar(str)-1)
    ```

    ```{r}
    # 4.
    # Write function bmi that calculates body mass index (bmi = weight / height2).
    # if bmi <= 18.5 return "Underweight"
    # if bmi <= 25.0 return "Normal"
    # if bmi <= 30.0 return "Overweight"
    # if bmi > 30 return "Obese"

    bmi <- function(weight, height) {
       var_bmi = weight / height^2
      if (var_bmi <= 18.5) {
        print("Underweight")
      } else if (var_bmi <= 25.0) {
        print("Normal")
      } else if (var_bmi <= 30.0) {
        print("Overweight")
      } else {
       print("Obese")
      }
    }

    bmi(80.0,1.80)
    ```

    ```{r}
    # 5.
    # Write a function that takes an array of words and smashes them together into a sentence and returns the sentence. You can ignore any need to sanitize words  or add punctuation, but you should add spaces between each word. Be careful, there shouldn't be a space at the beginning or the end of the sentence!

    smash <- function(words){
        print(paste(words))
    }

    ex <- c('hello', 'world', 'this', 'is', 'great') 
    smash(ex)

    # Issues with codewars:
    # Codewars solution is smash <- function(words) paste(words, collapse = " ")
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    check_type <- function(num){
      if(num %% 3 == 0 & num %% 5 == 0){
        return('fizzbuzz')
      }else if(num %% 5 == 0){
        return('buzz')
      }else{
          if(num %% 3 == 0){
            return('fizz')
          }
        }
      }

    check_type(5)
    check_type(21)
    check_type(30)
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    mario_stairs <- function(val){
      for(i in 1:val){
        print(paste(rep('X', i)))
      }
    }

    mario_stairs(4)
    ```

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.

    #def mario_stairs(val):
        #for i in range(1,val+1):
            #print('X' * i)

    #mario_stairs(4)
    ```
